Evaluation of a Cache-Oblivious Data Structure [draft-1]

### 2008-05-31 paper/paper.tex
Updates to the draft paper.

\textit{\small 9$^{th}$ Twente Student Conference on IT, Enschede, June, 2008}

@@

\section{Related Work}
Many data structures and algorithms have been analyzed in the cache-oblivious model. Several new data structures and algorithms have been developed that perform optimally in this mo\-del as well. Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.

@@

Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory. Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.

@@

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file.
Figure \ref{hash-table} shows a hash table (with an index size of four,
storing three values) and the way it is stored in consecutive memory.

@@

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5320 processor (1.86 GHz, 4~MB cache) and 8~GB of main
memory and 2~GB swap file. Although the system is a multiprocessor system
(and the E5320 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table.
The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the (fixed) size of the index.
We will use the hash table with an index size of 10,000,000 for further
experiments, as it seems to perform best even for a relatively small test case.

@@

\section{Future work}
[[Measure different data structures]]

[[Measure performance in low memory environments (using disk)]]

[[Mesure cache performance specifically]]

It should be noted that in our experiments only part of the memory
hierarchy was used (up to the use of main memory). This still involves several
levels of processor cache, but it is not a very deep hierarchy. With a deeper
memory hierarchy, with greater differences in access times between the levels,
the cache-efficient data structures (the B-tree and the Bender set) should
perform relatively better.

\section{Conclusion}
The experimental results clearly show that in terms of both memory usage and
execution time, the cache-oblivious data structure proposed by Bender, Duan,
Iacono and Wu is outperformed by traditional data structures in our test scenario.
The advantage of more cache-friendly behavior does not appear to be large enough
to compensate for the increased complexity of the data structure, which results in
high ermemory requirements and increased computational overhead.
If the data structure does have asymptotic performance benefits, then realistic
work loads on current hardware systems are not enough to reveal this.
These findings are consistent with earlier results, such as obtained by
Rahman, Cole and Raman, and Olsen and Skov (see Related Work above).

Of course, this does not prove definitely that cache-oblivious data structures
are entirely without merit. It should be noted that the experiments performed
are limited in scope: only a single cache-oblivious data structure has been
examined, in a single test scenario, on a single hardware
and software platform. It is possible that in a different setting,
different cache-oblivious data structures do compare favorably to their traditional
counterparts. However, since the observed differences in performance are fairly
large, it is unlikely that small changes are enough to close the performance gap.

@@

\section{Results}

>>>>

\textit{\small 9$^{th}$ Twente Student Conference on IT, Enschede, June 23$^{rd}$, 2008}

@@

\section{Related Work}
\label{sect-related-work}
Many data structures and algorithms have been analyzed in the cache-oblivious model. Several new data structures and algorithms have been developed that perform optimally in this mo\-del as well. Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting \cite{prokop1999coa}.

@@

Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize separate-chaining hash tables for cache efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory. Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.

@@

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file.
Figure \ref{fig-hash-table} shows a hash table (with an index size of four,
storing three values) and the way it is stored in consecutive memory.

@@

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5335 processor (2~GHz, 4~MB cache) and 8~GB of main
memory (no swap file). Although the system is a multiprocessor system
(and the E5335 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table.
The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the (fixed) size of the index.
Although for this case the hash table with an index size of 1,000,000 is
slightly faster, we will use an index size of 10,000,000 for further
experiments, as these cases require more values to be stored.

@@

\section{Future Work}
It should be noted that in our experiments only part of the memory
hierarchy was used (up to the use of main memory). This still involves several
levels of processor cache, but it is not a very deep hierarchy. In a deeper
memory hierarchy, with greater differences in access times between the levels,
the cache-efficient data structures (the B-tree and the Bender set) should
perform better. Specifically, using disk-based storage as the lowest level of
storage seems a logical extension of our research.

In our experiments we did not measure cache efficiency specifically; instead,
we measured total execution time only, which is affected by several different
factors of which cache efficiency is only one.
To better understand how different factors influence performance, it would
be interesting to measure cache efficiency seperately and reporton actual
cache hits and misses on different boundaries of the memory hierarchy.

Finally, we only evaluated a single data structure in a single test
environment (even though we used more than one model to perform experiments).
To draw more general conclusions about the practical merits of cache-oblivious
data structures, it will be necessary to perform experiments at a larger scale,
comparing multiple cache-oblivious data structures and using scenarios that
differ more.

\section{Conclusion}
The experimental results clearly show that the cache-oblivious data structure
proposed by Bender, Duan, Iacono and Wu is outperformed by traditional data
structures in our test scenario, in terms of both execution time and memory use.
The advantage of more cache-friendly behavior does not appear to be large enough
to compensate for the increased complexity of the data structure, which results in
higher memory requirements and increased computational overhead.
If the data structure does have asymptotic performance benefits, then realistic
work loads on current hardware systems are not enough to reveal this.
These findings are consistent with earlier results, such as obtained by
Rahman, Cole and Raman, and Olsen and Skov (see Section \ref{sect-related-work}).

Of course, this does not prove conclusively that cache-oblivious data structures
are entirely without merit. Our experiments are limited in scope: only a single
cache-oblivious data structure has been examined, in a single test scenario,
on a single hardware
and software platform.
It is possible that in a different setting, different cache-oblivious data
structures do compare favorably to their traditional counterparts.
However, since the observed differences in performance are fairly large,
it is unlikely that small changes are enough to close the performance gap.

@@
